---
title: "PROBABILITY AND STOCHASTIC PROCESSES CAT 1"
author: "Jeremy Gachanja"
date: "7/13/2022"
output: pdf_document
---

# 1. Peter is in jail and has 20 dollars; he can get out on bail if he has 64 dollars. A guard agrees to make a series of bets with him. If Smith bets A dollar, he wins A dollar with probability 0.55 and loses A dollar with probability 0.45. Find the probability that he wins 8 dollars before losing all of his money if

## a. he bets 5 dollars each time (timid strategy).

```{r}
i_1 = 5
n = 8
p=0.55
q=0.45
p_winning = (1-(q/p)^i_1)/(1-(q/p)^n)
print(p_winning)
```
From the above via the timid approach he has a 0.79 chance of wining his freedom

## b. he bets, each time, as much as possible but not more than necessary to
bring his fortune up to 64 dollars (bold strategy).

### solution

$$ P_{20} = 0.55P_{40}+0.45P_0$$

$$P_{40} = 0.55P_{64}+0.45P_{16}$$

$$P_{16} = 0.55P_{32}+0.45P_0$$

$$P_{32} = 0.55P_{64}+0.45P_0$$
The only known probabilities are $P_{64}, P_{0}$ which are 1 and 0 respectively. We can therefore estimate $P_{40}, P_{16}, P_{32}$ recursively from the equation (i) to (iv)
```{r}
p_32 = 0.55*(1)+0.45*(0)
p_16 = 0.55*(p_32)+0.45*(0)
p_40 = 0.55*(1)+0.45*(p_16)
p_20 = 0.55*(p_40)+0.45*(0)

print(p_20)
```
Therefore the probability of wining using the bold approach would be 0.3773

## c. Which strategy gives Smith the better chance of getting out of jail?

### solution

The timid appraoch gives him higher chances of wining his freedom with the probability of 0.79


# 2. Derive the formulas for the Gambler’s Ruin Problem for situation when p=q and not p=q . Also describe how to becoming infinitely rich or getting ruined in the same Gambler’s Ruin setting [Hint: Check several literatures on google and other text books]
$$\text{Deriving the formula for } p \neq q\\$$

$$\text{We start from the foundational equation that } q_z = pq_{z+1} = qq_{z-1}$$
$$\text{Here we can generalise that the solution to this equation can be found such that where A, B are real numbers then we can rewrite the above as: }\\ Aq_z + Bq_z\prime \rightarrow 1$$
$$\text {Recall that the above equation can be generalised to }\\ pP_i + qP_i = pP_{i+1} + qP_{i-1}$$
$$\text{Picking the like terms together we get }\\
pP_{i+1}-pP_i = qP_i - qP_{i-1}$$

$$p(P_{i+1}-P_i) = q(P_i-P_{i-1})$$
$$P_{i+1}-P_i = \frac{q}{p}(P_i-P_{i-1})$$

$$\text{In this case lets substitute i with numbers, such that i = 1. This would lead to: }$$
$$P_{2}-P_1 = \frac{q}{p}(P_1-P_{0})$$

$$\text{Recall that }P_0 = 0 \text{ As such the formula reduces to: }$$

$$P_{2}-P_1 = \frac{q}{p}(P_1)$$

$$\text{Consequently we can expand this to i = 2, such that: }$$

$$P_{3}-P_2 = \frac{q}{p}(P_2-P_{1})\ \text{Recall that }P_2-P_1 = \frac{q}{p}(P_1)\\$$
$$\text{ This would thus lead to }P_3 -P_2 = \frac{q}{p}(\frac{q}{p}(P_1)) \rightarrow P_3 - P_2 = (\frac{q}{p})^2P_1$$

$$\text{This can be generalised to} P_{i+1}-P_i = (\frac{q}{p})^i(P_{i+1}-P_{i-1})\\$$
$$\text{Given that }q_0 = 1 \text{ and } q_a = 0 \text{ then 1 will be come for i = 0 }$$

$$A+B = 1$$
$$\text{for i = a, 1 will become}$$

$$B(\frac{q}{p})^a = q_a$$



# 3. Tom starts with 5, and p = 0.63: What is the probability that:

## a.Tom obtains a fortune of N = 12 without going broke? 

### solution

```{r}
n = 12
p=0.63
q=1-p
i = 5
p_win = (1-(q/p)^i)/(1-(q/p)^n)

print(p_win)
```


## b. What is the probability that Tom will become infinitely rich? 

### solution

```{r}
p_infinite_w = 1-(q/p)^5
print(p_infinite_w)
```
From the above output, the chances of becoming infinitely rich are 0.93

## c.If Tom instead started with i = 2, what is the probability that he would go broke?

```{r}
n = 12
p=0.63
q=1-p
i = 2
p_win = (1-(q/p)^i)/(1-(q/p)^n)

print(1-p_win)
```
From the above output, the chances of being ruined are 0.34


# 4. Collins bought a share of stock for 10, and it is believed that the stock price moves (day by day) as a simple random walk with p = 0.6. What is the probability that Collins’ stock reaches the high value of $25 before the low value of 4?

```{r}
p = 0.6
q=1-p
i=10
n=25
a=n-i
b=4
p_stock_movement = (1-(q/p)^b)/(1-(q/p)^(a+b))
print(p_stock_movement)

```
From the output above, the chances of the stock going up to 25 before going down to 4 is 0.802
